---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
``` {r loading, echo = TRUE}
library("knitr")
library("rmarkdown")
## Constants ##
localFile <- "activity.zip"

## Download and unzip 
# Skip unzipping if data exist
if (!file.exists("activity.csv")) {
  unzip(localFile)
}

## load data
activitydata <- read.csv("activity.csv")
## select complete cases 
actcompltedata <-activitydata[complete.cases(activitydata),]
```



## What is mean total number of steps taken per day?
``` {r plot, echo = TRUE}
library(ggplot2) 
tabledata <- aggregate(steps ~ date,actcompltedata, sum)

myplot <- qplot(tabledata$steps, geom="histogram", binwidth = 5000) 
myplot + labs(list(title = "Total Steps Per Day", x = "Steps", y = "Frequency"))

mean(tabledata$steps)
median(tabledata$steps)
```

The mean number of steps taken per day is `r as.integer(mean(tabledata$steps),0)`.
The median number of steps taken per day is `r median(tabledata$steps)`.

## What is the average daily activity pattern?

```{r}
# preprocessing data for plot
stepsinterval <- aggregate(steps ~ interval, actcompltedata, mean)

intervalplot <- qplot(stepsinterval$interval, stepsinterval$steps, geom="line")
intervalplot + labs(list(title = "Average number of steps over all days", x = "Interval", y = "Average number of steps"))

#preparefuction 
# find row with max of steps
maxsteps <- which.max(stepsinterval$steps)
# find interval with this max
stepsinterval[maxsteps, ]
pit <- stepsinterval[maxsteps, ]

```
The interval `r pit[1]`, contains the maximum number of steps (`r round(pit[2],1)`).

## Imputing missing values

```{r}
sum(is.na(activitydata))
```

There are `r sum(is.na(activitydata))` missing values in the dataset. 

For imputing missing values I use mean for that 5-minute interval caluclated earlier and presented on above graph.
```{r}
dataimputed <- activitydata
for (i in 1:nrow(dataimputed)) {
  if (is.na(dataimputed$steps[i])) {
    intervalvalue <- dataimputed$interval[i]
    stepsvalue <- stepsinterval[
      stepsinterval$interval == intervalvalue,]
    dataimputed$steps[i] <- stepsvalue$steps
  }
}

tabledataimputed <- aggregate(steps ~ date,dataimputed, sum)

myplot <- qplot(tabledataimputed$steps, geom="histogram", binwidth = 5000) 
myplot + labs(list(title = "Total Steps Per Day", x = "Steps", y = "Frequency"))

mean(tabledataimputed$steps)
median(tabledataimputed$steps)
```

The mean number of steps taken per day is `r as.integer(mean(tabledataimputed$steps),0)`.
The median number of steps taken per day is `r as.integer(median(tabledataimputed$steps),0)`.

The mean with imputed values remains the same as mean calculated with NA values removed. Median is now higher by one step. 

## Are there differences in activity patterns between weekdays and weekends?
```{r}
df <- dataimputed ## save typing long name 
# convert date to Date class
df$date <- as.Date(df$date, "%Y-%m-%d")
# add a new column for storing day of the week 
df$day <- weekdays(df$date)
# add a new column to determine weekend or not
df$day_type <- c("weekday")
# If day is Saturday or Sunday, make weekend
for (i in 1:nrow(df)){
  if (df$day[i] == "Saturday" || df$day[i] == "Sunday"){
    df$day_type[i] <- "weekend"
  }
}

df_imputed_steps_by_interval <- aggregate(steps ~ interval + day_type, df, mean)

## plot it 
qplot(interval, 
      steps, 
      data = df_imputed_steps_by_interval, 
      type = 'l', 
      geom=c("line"),
      xlab = "Interval", 
      ylab = "Number of steps", 
      main = "") +
  facet_wrap(~ day_type, ncol = 1)
```

```{r cleanup}
# remove the data frames to free memory
rm(localFile, activitydata, actcompltedata,tabledata,myplot, stepsinterval,maxsteps, pit, dataimputed, tabledataimputed, df,df_imputed_steps_by_interval)
```
